---
title: "21MAP500 Coursework"
date: "Last updated: `r format(Sys.time(), '%m, %d, %Y')`"

output: 
  html_document:
    toc: true 
    number_sections: true 
---

```{r,  results='hold', echo = FALSE}


#install.packages("janitor")

library("tidyverse")   #call out the packages we need
library("here")
library("lubridate")
library("ggrepel")
library("gridExtra")
library("RColorBrewer")
library("readxl")
library("janitor")




```


#Question 1
##1.a
```{r, results='hold'}

read_lines(file = here("data", "nasa_global_temperature.txt"), n_max = 5L) #read the file first to see what kind of data we are using 

read_table(here("data", "nasa_global_temperature.txt"), skip = 3) -> nasa_temp #read the file into a tibble dataframe and save into a new  variable

nasa_temp %>% 
  slice(2:142) %>% #get rid of the first row
  select(Year, No_Smoothing) %>% #select the columns we need
  mutate(Year = parse_date(Year, format = "%Y")) %>%  #change type of "Year" to date type
  rename(date = Year, temp = No_Smoothing) -> nasa_temp  #rename "Year" to "date", "No_Smoothing" to "temp" and save to variable nasa_temp

nasa_temp %>% 
  ggplot(mapping = aes(x = date, y = temp, colour = temp)) +
  geom_line(size = 1) +  #use "date" as x axis, "temp" as y axis to plot a line graph, change line size
  labs(    #change titles
    title = "Change of global average temperature from 1880-2020",
    x = "Year 1880-2020",
    y = "Average temperature across global land and ocean surfaces in °C",
    caption = "Global average temperature has became higher than before."
  ) +
  scale_color_continuous(low = "blue", high = "red") +  #change colour to emphasis temperature rise
  guides(colour = FALSE) +  #remove redundant legend 
  theme_classic()+  #remove background
  theme(
        axis.line.x = element_line(size = 1),      #adjust x axis size 
        axis.line.y = element_line(size = 1),      #adjust y axis size 
        axis.title.x = element_text(size = 12),    #adjust x axis title size  
        axis.title.y = element_text(size = 10.5)   #adjust y axis title size 
        ) 

```


##1.b
```{r,  results='hold'}

read_lines(file = here("data", "nasa_arctic_sea_ice.csv"), n_max = 10)  #read the file first to see what kind of data we are using 

nasa_ice <- read_csv2(file = here("data", "nasa_arctic_sea_ice.csv"),
          col_select=c(year, extent)) #read the file into a tibble data frame and save into a new variable

          

nasa_ice %>% 
  mutate(across(year, as.character)) %>%  #change "year" into character type
  mutate(year = parse_date(year, format = "%Y")) %>%  #then to date type
  rename(date = year, ice = extent) -> nasa_ice  #rename columns and save the result to variable nasa_ice

nasa_ice %>% 
  ggplot(mapping = aes(x = date, y = ice, size = ice)) +
  geom_point(alpha = 0.3) -> p #use "date" as x axis, "ice" as y axis to plot a scatter plot

p +
  labs(    #change titles
    title = "Change of minimum arctic sea ice from 1979-2020",
    x = "Year",
    y = "Minimum arctic sea ice extent in million square km",
    caption = "The Arctic ice started decline around 1995"
  ) +
  theme_classic() + #remove background
  theme(
        axis.line.x = element_line(size = 1),     #adjust x axis size 
        axis.line.y = element_line(size = 1),     #adjust y axis size 
        axis.title.x = element_text(size = 12),   #adjust x axis title size  
        axis.title.y = element_text(size = 12)    #adjust y axis title size  
        ) +
  guides(size = FALSE)  #remove legend




```


##1.c
```{r,  results='hold'}

read_lines(file = here("data", "nasa_sea_level.csv"), n_max = 10)  #read the file first to see what kind of data we are using 


nasa_sea <- read_table(file = here("data", "nasa_sea_level.csv"), skip = 48, col_names = FALSE)  #read the file into a tibble skip  the first 48 lines we don't need and save into a new variable


nasa_sea %>% 
  select(X3, X12) %>%  #select the column we need
  rename(date = X3, sea = X12) %>% #rename the column to "date" and "sea"
  mutate(date = date_decimal(date, tz = "UTC")) -> nasa_sea #change the type to date type and save to variable nasa_sea

nasa_sea %>% 
  ggplot(mapping = aes(x = date, y = sea, colour = sea)) +
  geom_col() -> p #use "date" as x axis, "sea" as y axis to plot a histogram


p +
  labs(    #change titles
    title = "Change of sea level between 1993-2021 compared to a reference year",
    x = "Year",
    y = "Change in sea level compared to a reference year in mm",
    caption = "Around 2005, sea level started to exceed the reference year sea level."
  ) +
  scale_color_continuous(low = "tan3", high = "skyblue") + #change colour to emphasis sea level risen 
  theme_classic() + #remove background
  theme(
        axis.line.x = element_line(size = 1),     #adjust x axis size 
        axis.line.y = element_line(size = 1),     #adjust y axis size 
        axis.title.x = element_text(size = 12),   #adjust x axis title size  
        axis.title.y = element_text(size = 12),   #adjust y axis title size  
        panel.grid.major.x = element_blank(),     # remove thick vertical grid lines
        panel.grid.minor.x = element_blank()      # remove thin vertical grid lines
        ) +
  guides(colour = FALSE)  #remove legend
  
  
```


##1.d
```{r, results='hold'}

read_lines(file = here("data", "nasa_carbon_dioxide.txt"), n_max = 10) #read the file first to see what kind of data we are using 

nasa_co2 <- read_table(file = here("data", "nasa_carbon_dioxide.txt"), 
                     skip = 52,) #read the file into a tibble and save into a new variable
nasa_co2 %>% 
  select(average, alized) %>%  #select the columns we need
  rename(date = average, co2 = alized) %>%  #rename the columns to "date" and "co2"
  mutate(date = date_decimal(date)) %>%   #use decimal_date to calculate the fraction year
  mutate(date = as_date(date)) %>% #change to date type
  mutate(date = floor_date(date, unit = "month")) %>%  #round date to the first day of the month
  filter(co2 > 0) -> nasa_co2 # filter out outliers 
  
  
nasa_co2 %>% 
  ggplot(mapping = aes(x = date, y = co2, colour = co2)) + 
  geom_line(size = 1) -> p #visualize data with a scatter plot, "date" as x axis, "co2" as y axis

                     
p +
  labs(    #change titles
    title = "Average global CO2 level from 1958 to 2021",
    x = "Year",
    y = "Average global CO2 level in parts per million (ppm)",
    caption = "The global co2 level has increased dramatcally since 1958."
  ) +
  scale_color_continuous(low = "skyblue", high = "deeppink") + #change colour to emphasis co2 risen 
  theme_classic() + #remove background
  theme(
        axis.line.x = element_line(size = 1),     #adjust x axis size 
        axis.line.y = element_line(size = 1),     #adjust y axis size 
        axis.title.x = element_text(size = 12),   #adjust x axis title size 
        axis.title.y = element_text(size = 12),   #adjust y axis title size  
        panel.grid.major.x = element_blank(),     #remove thick vertical grid lines
        panel.grid.minor.x = element_blank()      #remove thin vertical grid lines
        ) +
  guides(colour = FALSE)  #remove legend
  
                     



```


##1.e
```{r,  results='hold'}


nasa_sea %>% 
  mutate(date = as_date(date)) ->nasa_sea #convert nasa_sea$date into date type instead of datetime type to be join with other later


nasa_temp %>% 
  full_join(nasa_ice, by = c("date" = "date"), keep = FALSE) %>% 
  full_join(nasa_sea, by = c("date" = "date"), keep = FALSE) %>%
  full_join(nasa_co2, by = c("date" = "date"), keep = FALSE) -> nasa #join all the tibble together and save in to "nasa" 


nasa

```


##1.f
```{r,  results = 'hold'}

nasa %>% 
  select(date,co2, temp) %>%  #select only the columns we need
  filter(date >= "1960-01-01" & date <= "2020-01-01") %>% # filter out the time frame we want 
  ggplot(mapping = aes(x = co2, y = temp, colour = date)) +  
  geom_point(size = 3)  + #use "co2" as x axis, "temp" as y axis, "date for colour" to visualize the scatterplot, and use use size to change the point size
  labs(title = "Global surface tempature VS Global CO2 levels in year 1960-2020",
        x = "Average global CO2 level in parts per million (ppm)",
        y = "Average tempature across global surfaces in °C", 
        caption = "Higher co2 level has increased global average temperature within recent years.",
        colour =  "Past 60 Years" 
  ) + # use labs() to change the titles and axis name
  theme_classic() +  #remove background
  theme(
        axis.line.x = element_line(size = 1),     #adjust x axis size 
        axis.line.y = element_line(size = 1),     #adjust y axis size 
        axis.title.x = element_text(size = 12),   #adjust x axis title size 
        axis.title.y = element_text(size = 12)
      )
  


```




#2
##2.a
```{r,  results='hold'}

read_table(here("data", "luthi_carbon_dioxide.txt"), skip = 773) %>%  #read in the file and skip the first 773 lines
  rename(CO2 = `CO2(ppmv)`) %>% #rename column name
  rename(yrbp = `Age(yrBP)`) -> historic_co2  # rename column name and save to new variable

historic_co2 
```




##2.b
```{r,  results='hold'}

historic_co2 %>% 
  mutate(yrbp = yrbp + 13) -> historic_co2_modified  #mutate the yrbp reference year to 2021 
  

nasa_co2 %>% 
  mutate(date = year(date)) %>% # retrieve the year of the date
  mutate(yrbp = 2021 - date) %>%  # mutate the yrbp reference year to 2021
  select(yrbp, co2) %>%  #select the columns we need 
  group_by(yrbp) %>%  #group yrbp for calculation
  summarise(co2 = mean(co2, na.rm = TRUE)) %>%  #average the "co2" column
  full_join(historic_co2_modified, by = c("yrbp" = "yrbp", "co2" = "CO2")) -> combined_co2  #join nasa_co2 tibble and historic_co2_modified tibble together as requested and save in to new variable combined_co2
 
combined_co2


  


```



##2.c
```{r,  results='hold'}

combined_co2 %>% 
  ggplot(mapping = aes(x = yrbp, y = co2)) +  #set "yrbp as x axis and "co2" as y axis as the question 
  geom_line(size = 1) +  #choose geom_line to create line graph and set line size as 1 to make it thicker
  scale_x_reverse(breaks = c(800000, 600000, 400000, 200000, 0),  #reverse the x axis value and setting the x axis breaks from 800,000 to 0 
                  labels = c("800,000", "600,000", "400,000", "200,000", "0")  #change the scientific notation to the label we want
                  ) +
  labs(x = "Years before present",  #change x axis title
       y = "Carbon dioxide [ppm]",  #change y axis title
       ) +
  theme_classic() + #change the theme to classic to remove the background grid
  theme(axis.title.x = element_text(size = 14), #adjust x axis title size
        axis.title.y = element_text(size = 14), #adjust y axis title size
        axis.line.x = element_line(size = 1),   #adjust x axis line size
        axis.line.y = element_line(size = 1)    #adjust y axis line size
        ) -> p # save to variable p


  
point_coords <- c(combined_co2$yrbp[1], combined_co2$co2[1])  #set the start point coordinate
label_coords <- point_coords + c(60000, -7)   #set the label coordinate





#curve annotation
p_curve <-  #save the curve annotation to p_curve variable
  annotate(
    geom = "curve",  #create the curve line
    x = label_coords[1],  #set the start point coordinate for x
    y = label_coords[2],  #set the start point coordinate for y
    xend = point_coords[1],   #set the end point coordinate for x
    yend = point_coords[2],   #set the end point coordinate for y
    curvature = -0.4,   #set the curvature
    arrow = arrow(length = unit(2.5, "mm")),   #add arrow on the start point set size
    colour = "salmon",   #set colour to "salmon"
    size = 1    #set line size
  )
  
  
  
# Plot curve and point annotation with explanatory text  
p +  
  p_curve +  #combine the plot and curve annotation and annotation text together
  annotate(
    geom = "text",  #set annotation text
    x = label_coords[1] + 100000,  #set text coordinate
    y = label_coords[2],  #set y text coordinate
    label = c("2021 average:\n 417 ppm"),  #set the label content
    hjust = 0.5,  #adjust the horizontal position to the middle
    vjust = 1,    #adjust the vertical position 
    lineheight = 1,  #adjust the line height in between lines
    colour = "salmon",  #set colour
    size = (5)  #set font size
    )



  





```



#3
##3.a
```{r,  results='hold'}

#install.packages("readxl")



read_lines(file = here("data", "nsidc_sea_ice_daily_extent.xlsx"), n_max = 10)  #read the file first to understand what data type it is

read_xlsx(here("data", "nsidc_sea_ice_daily_extent.xlsx")) -> sea_ice_1   #read in the file as tibble and save to new variable

sea_ice_1 %>% 
  select(!47:49)%>%  #remove the column we do not need
  rename(month = ...1, day = ...2) %>%   #rename the first 2 columns
  fill(month) %>%   #fill up the missing value by default direction will go down
  pivot_longer(cols = 3:46, names_to = c("year"), values_to = "extent") %>%   #pivot the tibble longer by making the years columns in to one column, name "year" and the values to "extent"
  mutate(year = parse_date(year, format = "%Y")) %>%   #change the "year" column to date type
  mutate(year = year(year)) %>%          #retrieve only the year part of the date
  mutate(year = as.integer(year)) %>%    #change in to integer type as requested
  mutate(day = as.integer(day)) %>%      #change in to integer type as requested
  #mutate(month = as_date(month, format = "%B"))   #does not work
  mutate(month = case_when(      #manually changing the month to integer 
    month == "January" ~ 1L,
    month == "February" ~ 2L,
    month == "March" ~ 3L,
    month == "April" ~ 4L,
    month == "May" ~ 5L,
    month == "June" ~ 6L,
    month == "July" ~ 7L,
    month == "August" ~ 8L,
    month == "September" ~ 9L,
    month == "October" ~ 10L,
    month == "November" ~ 11L,
    month == "December" ~ 12L,
  )) -> sea_ice_1  
  

sea_ice_1  


```


##3.b
```{r,  results='hold'}

sea_ice_1 %>% 
  group_by(month, year) %>%  #group by month and year 
  summarise(avg_month_extent = mean(extent, na.rm = TRUE)) -> monthly_avg_extent   #calculate the average value of extent column and save in to new variable

sea_ice_1 %>% 
  group_by(month, year) %>%   #group by month and year
  filter(year == 1979) %>%    #filter to get value in 1979
  summarise(month_specific_baseline_extent = mean(extent, na.rm = TRUE)) %>%   #calculate the average of "extent" in 1979  
  select(!year) -> baseline  #select all the columns except for "year" 

  
monthly_avg_extent %>%   
  left_join(baseline, by = c("month" = "month")) %>%    #join two tibble together 
  mutate(proportion_baseline_extent = avg_month_extent / month_specific_baseline_extent) %>%   #calculate the "proportion baseline extent" as instructed 
  select(year, month, proportion_baseline_extent) -> sea_ice_2  #select the columns we need and save it to new variable 
  
sea_ice_2

```


##3.c
```{r,  results='hold'}


sea_ice_2 %>% 
  ggplot(mapping = aes(x = year, y = month, fill = proportion_baseline_extent)) +  #use "year" as x axis, "month" as y axis and  "proportion_baseline_extent" as fill value
  geom_tile() -> p   #to create a tile graph and save to p

p + scale_x_continuous(expand = expansion(add = c(-1,-1))) +     #remove padding
  scale_y_continuous(expand = expansion(add = c(0,0)),           #remove padding
                     breaks = c(1:12), labels = c("January",     #set the x axis breaks and labels
                                                  "February",
                                                  "March",
                                                  "April",
                                                  "May",
                                                  "June",
                                                  "July",
                                                  "August",
                                                  "September",
                                                  "October",
                                                  "November",
                                                  "December"
                                                  )) +  
  
  labs(title = "Sea ice (northern hemisphere)",   #create title
    x = "Year",                                   #create x axis title
    y = "Month",                                  #create y axis title
    fill = "Proportion of 1979 extent"            #create legend axis title
    ) +
  scale_fill_distiller(palette = "RdPu",          #change the fill value color
                       breaks = c(0.6, 0.7, 0.8, 0.9, 1.0),   #set legend break
                       limits = c(0.5, 1.1)                   #set legend limit
                       ) +
  theme_classic() +   #set the theme same as question
  theme(plot.title = element_text(hjust = 0.5,        #adjust title position
                                  size = rel(1.5)),
        axis.line.x = element_line(size = 1),         #adjust x axis size 
        axis.line.y = element_line(size = 1),         #adjust y axis size 
        axis.title.x = element_text(size = 12, face = "bold"),   #adjust x axis title size and font 
        axis.title.y = element_text(size = 12, face = "bold")    #adjust y axis title size and font 
        )
  


```


#4
##4.a
```{r,  results='hold'}

read_lines(file = here("data", "stop_and_search.csv"), n_max = 10)   #read the file first to understand what data it is

stop_search_1_raw <- read_csv(here("data", "stop_and_search.csv"))  #read in the file as tibble and save to new variable


stop_search_1_raw %>% 
  clean_names() %>%   #change column names to snake case
  rename(stops = number_of_stop_and_searches,   #rename columns
         population = population_by_ethnicity, 
         rate = rate_per_1_000_population_by_ethnicity
          ) %>% 
  mutate(stops = parse_number(stops, na = c("", "NA")),   #change the "stops" type to integer
         population = parse_number(population, na = c("", "NA")),   #change the "population" type to integer
         rate = parse_number(rate, na = c("", "NA"))    #change the "rate" type to integer
          ) %>% 
  filter(ethnicity == "All" |        #filter out the ethnicity we need
          ethnicity == "Asian" |
          ethnicity == "Black" |
          ethnicity == "White" |
          ethnicity == "Other"  
           ) %>% 
  rename(year = time) %>%  #rename "time" to "year"
  select(year, ethnicity, legislation_type, geography, stops, population, rate) -> stop_search_1   #select the columns we need and save to new variable  

stop_search_1

```






##4.b
```{r,  results='hold'}


  

stop_search_1 %>% 
  group_by(year, ethnicity, legislation_type, geography) %>%   #group by the columns value to create new tibble that will be join later
  summarise(rate_white = mean(rate, na.rm = TRUE)) %>%   #calculate the average rate
  filter(ethnicity == "White") -> avg_rate_white    #filter out only the white ethnicity and save it to new variable


  
stop_search_1 %>% 
  left_join(avg_rate_white, by = c("year" = "year", "legislation_type" = "legislation_type", "geography" =  "geography")) %>%  #join the "avg_rate_white" to "stop_search_1"
  rename(ethnicity = ethnicity.x) %>%   #rename column 
  mutate(relative_disparity = rate / rate_white) %>%   #calculate "relative_disparity" as instructed and create new column to save value
  select(!ethnicity.y) -> stop_search_2   #select only the columns we need and save to new variable
  

stop_search_2 



```



##4.c & d


###Question 1:  
Does black ethnicity has higher chance to be stop and search by police in general?

```{r,  results='hold'}







stop_search_2 %>% 
  group_by(ethnicity) %>%   #grouped by same ethnicity
  mutate(ethnicity_rate = mean(rate, na.rm = TRUE)) %>%   #average the rate between different ethnicity
  select(ethnicity, ethnicity_rate) %>%    #select the column we need
  ggplot(mapping = aes(x = ethnicity_rate, y = ethnicity, fill = ethnicity)) +    #set x axis and y axis and fill with ethnicity
  geom_col(width = 0.5) -> p   #plot the graph and save to variable p

p +
  theme_classic() +
  theme(
    axis.line.x = element_line(size = 1),  #adjust x axis line size      
    axis.line.y = element_line(size = 1),  #adjust y axis line size 
    axis.title.x = element_text(size = 12, face = "bold"),   #adjust x axis title size and font
    axis.title.y = element_text(size = 12, face = "bold")    #adjust y axis title size and font
        ) +
  labs(
       title = "Total stops and searches rate between different ethnicity",   #change titles, axes labels and caption
       x = "Total stops and searches rate",
       y = "Ethnicity",
       caption = "Black ethnicity has much higher chance to be stop and search by police."
      ) +
  scale_x_continuous(
    expand = expansion(add = c(0, 5))) +  #remove padding
  scale_fill_manual(values = c("springgreen", "tan1", "gray23", "khaki", "skyblue")) +   #change the colour of ethnicity in legend
  guides(fill = FALSE)  #remove legend




```


###Question 2
What are the stops and searches rate difference between Black, Asian and White ethnicity in three most populated area?  


```{r, fig.width=10, fig.height=6, results='hold'}



stop_search_2 %>% 
  mutate(year = as_date(year, format = "%Y/%y")) %>%   #change the character type to date type 
  mutate(year = year(year)) %>%  #retrieve only the year
  filter(geography != "All - excluding BTP" &   #excluding area in "all"  
        geography !="All - excluding BTP and Greater Manchester" &
        geography != "All - including BTP" &
        geography != "All - including BTP and excluding Greater Manchester") %>% 
  arrange(desc(population)) %>%   #reordered the population to find out most populated area
  filter(geography == "Metropolitan Police" |    #filter out the area we need
         geography == "West Midlands" |
         geography == "Greater Manchester"  
        ) %>% 
  select(year, ethnicity, stops, rate, geography) -> top_3_populated_df   #select the columns we need and save to new variable

top_3_populated_df %>% 
  filter(ethnicity %in% c("Black", "Asian", "White")) %>%   #filter out the ethnicity we want to observe  
  ggplot(mapping = aes(x = year,  y = rate, fill = ethnicity)) +  #set the axes to plot graph
  geom_col(position = "dodge") +    #set position "dodge" to avoid stacking 
  facet_wrap(facets = vars(geography)) + #use facet function to plot different graph in different area
  labs(
       title = "Stops and searches rate between different ethnicity in most populated area from 2006-2020",   #change titles and caption
       x = "Year",
       y = "Stops and searches rate by police",
       caption = "In all three area black ethnicity has the highest chance to be stop and search by police, where white ethnicity has the lowest,"
      ) +
  scale_fill_manual(values = c("tan1", "gray23", "skyblue")) +    #change the colour of ethnicity in legend
  theme_classic() +    #remove background
  theme(
    axis.line.x = element_line(size = 1),  #adjust x axis line size      
    axis.line.y = element_line(size = 1),  #adjust y axis line size 
    axis.title.x = element_text(size = 12, face = "bold"),   #adjust x axis title size and font
    axis.title.y = element_text(size = 12, face = "bold")    #adjust y axis title size and font
        ) +
  scale_y_continuous(
    expand = expansion(add = c(0, 5)))   #remove padding
  


```
My thought was that more populated area should have more multi-culture experiences and the police will be less bias than other area, 
but the result shows that other ethnicity still have much higher chance to be stop and search than white ethnicity. 



###Question 3
What is the rate for Asian ethnicity to be stop and search compare to white ethnicity?


```{r, results='hold'}

stop_search_2 %>% 
  mutate(year = as_date(year, format = "%Y/%y")) %>%   #change the character type to date type 
  mutate(year = year(year)) %>%   #retrieve only the year
  filter(ethnicity == "Asian") %>%   #filter out Asian
  group_by(year) %>%   #grouped by year for calculation
  mutate(asian = mean(rate, na.rm = TRUE)) %>%   #calculate average rate for Asian 
  mutate(white = mean(rate_white, na.rm = TRUE)) %>%  #calculate average rate for White 
  select(year, asian, white) -> df  #select columns we need save to new variable
  
df %>% 
  pivot_longer(cols = 2:3, names_to = "ethnicity", values_to = "rate") %>%   #transform tibble to make ethnicity to one column 
  ggplot(mapping = aes(x = year, y = rate, colour = ethnicity)) +  #create line graph include Asian and white stop and search rate 
  geom_line(size = 2) -> p   #make line thicker
  
  
p + 
  labs(
       title = "Stops and searches rate between Asian and White ethnicity \n from 2006-2020",   #change titles, axes labels and caption
       x = "Years",
       y = "Stops and searches rate by police",
       caption = "Asian ethnicity has a higher rate to be stop and search by the police than White ethnicity.",
       colour = "Ethnicity"
      ) +
  theme_classic() +    #remove background
  theme(
    axis.line.x = element_line(size = 1),  #adjust x axis line size      
    axis.line.y = element_line(size = 1),  #adjust y axis line size 
    axis.title.x = element_text(size = 12, face = "bold"),   #adjust x axis title size and font
    axis.title.y = element_text(size = 12, face = "bold")    #adjust y axis title size and font
        ) 
  
  

```











